---
title: "code"
output:
  pdf_document: default
  html_document: default
date: "2024-11-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(sf) # For spatial data if using maps
```

## Research Question

This study investigates how stress from price changes (PRICESTRESS) varies across different demographic groups, specifically focusing on age, race, and marital status.

## Methods and Data

The analysis uses the 2023 Household Pulse Survey dataset, which includes responses from individuals across the U.S. The variable PRICESTRESS indicates stress levels from price changes, with higher values indicating more stress. Demographic variables used in this analysis include:

- Age: Categorized into age groups (e.g., 18-29, 30-44, 45-64, 65+).
- Race: Self-reported racial categories.
- Marital Status: Categories like married, single, widowed, etc.
- Weights (PWEIGHT) were applied to ensure representativeness. Descriptive and inferential statistics, along with visualizations, were used to identify and quantify differences.
```{r}
files <- list.files("data", recursive = TRUE, full.names = TRUE, pattern = "puf")
pulse <- map_df(files, read_csv)

# Prepare relevant variables
pulse_clean <- pulse %>%
  select(PRICESTRESS, TBIRTH_YEAR, RRACE, MS, PWEIGHT, EST_ST, REGION) %>%
  filter(!is.na(PRICESTRESS), !is.na(TBIRTH_YEAR), !is.na(RRACE), !is.na(MS))
```

## Analysis
### Descriptive Statistics
```{r}
# Create AGE and age_group variables
pulse_data <- pulse_clean %>%
  mutate(
    AGE = 2024 - TBIRTH_YEAR, # Calculate AGE from TBIRTH_YEAR
    age_group = case_when(
      AGE < 30 ~ "18-29",
      AGE < 45 ~ "30-44",
      AGE < 65 ~ "45-64",
      TRUE ~ "65+"
    )
  )

# Map RRACE (race) and MS (marital status) to meaningful labels
pulse_data <- pulse_data %>%
  mutate(
    RRACE = case_when(
      RRACE == 1 ~ "White, Alone",
      RRACE == 2 ~ "Black, Alone",
      RRACE == 3 ~ "Asian, Alone",
      RRACE == 4 ~ "Other Race/Combination",
      TRUE ~ "Unknown" # For any unexpected value
    ),
    MS = case_when(
      MS == 1 ~ "Now married",
      MS == 2 ~ "Widowed",
      MS == 3 ~ "Divorced",
      MS == 4 ~ "Separated",
      MS == 5 ~ "Never married",
      MS == -99 ~ "Not selected",
      MS == -88 ~ "Missing/Not reported",
      TRUE ~ "Unknown" # For any unexpected value
    )
  )

pulse_data <- pulse_data %>%
  filter(!PRICESTRESS %in% c(-99, -88))
```

## Histogram for numeric variables
```{r}
# Select numeric columns from pulse_data
numeric_pulse_data <- pulse_clean %>%
  select(where(is.numeric)) # Select only numeric columns

# Pivot the data to a long format for plotting
pulse_long <- numeric_pulse_data %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Plot histograms for all variables using facets
ggplot(pulse_long, aes(x = value)) +
  geom_histogram(bins = 30, fill = "blue", color = "black", alpha = 0.7) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  theme_minimal() +
  labs(
    title = "Histograms of All Numeric Variables in pulse_data",
    x = "Value",
    y = "Frequency"
  )
```


```{r}
# Group data by age group
age_count_st <- pulse_data %>%
  group_by(age_group, EST_ST, REGION) %>%
  summarise(count=sum(PWEIGHT)) %>%
  ungroup() %>%
  mutate(REGION = as.character(REGION)) # Ensure REGION is a character

# Display the results
age_count_st

# Plot stress levels by age group across states
ggplot(age_count_st, aes(x = count,
                         y = reorder(EST_ST, count),
                         fill = REGION)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ age_group, scales = "free") +
  xlab("Total Count of Weighted Responses") +
  ylab("State FIPS") +
  ggtitle("Weighted Response Counts by Age Group, State, and Region") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3", name = "Region")

```

### Inferential Statistics
We conduct an ANOVA test to examine whether differences in stress levels are statistically significant across age groups, racial groups, and marital statuses.
```{r}
# ANOVA for age group
anova_age <- aov(PRICESTRESS ~ age_group, data = pulse_data)
summary(anova_age)

# ANOVA for race
anova_race <- aov(PRICESTRESS ~ RRACE, data = pulse_data)
summary(anova_race)

# ANOVA for marital status
anova_ms <- aov(PRICESTRESS ~ MS, data = pulse_data)
summary(anova_ms)

```
# Univariate Model
```{r}
pulse_clean <- pulse_clean %>%
  mutate(
    AGE = 2024 - TBIRTH_YEAR, # Calculate AGE from TBIRTH_YEAR
    age_group = case_when(
      AGE < 30 ~ "18-29",
      AGE < 45 ~ "30-44",
      AGE < 65 ~ "45-64",
      TRUE ~ "65+"
    )
  )

# Create univariate models for each predictor
model_age <- lm(PRICESTRESS ~ age_group, data = pulse_clean)
model_race <- lm(PRICESTRESS ~ RRACE, data = pulse_clean)
model_marital <- lm(PRICESTRESS ~ MS, data = pulse_clean)
model_region <- lm(PRICESTRESS ~ REGION, data = pulse_clean)

# Display summaries for each model
summary(model_age)   # Effect of age group on stress levels
summary(model_race)  # Effect of race on stress levels
summary(model_marital) # Effect of marital status on stress levels
summary(model_region) # Effect of region on stress levels

```



# Multivariate Model
```{r}
# Linear regression model
stress_model <- lm(
  PRICESTRESS ~ age_group + RRACE + MS + REGION,
  data = pulse_clean
)

# Summary of the model
summary(stress_model)
```

